 

### 2021-04-06 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. Foliage density mediates light regime, where denser upper canopies exposed to greater PAR values limit light to canopy interiors, lower canopy layers, and the understory. 
This light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @niinemetsPackingPhotosyntheticMachinery2009; @poorterMetaanalysisPlantResponses2019b].

@@

The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower daytime maximum temperature and vapor pressure deficit ($VPD$), and warmer higher nighttime minimum temperature in the understory, respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021, @davis_microclimatic_2019, @zellweger_seasonal_2019]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$ [S. R. Hardwick et al. 2015, Junker et al. 2018]. Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a]. These pine forests have also been observed to maintain warmer nighttime understory temperatures than above canopy, due to radiative heat loss from the soil, caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature. However, beyond a certain height, temperature and VPD offset plateaus [Junker et al. 2018]. At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].

@@

*Topography and altitude contribute to the heterogeneity of forest light environment, and thus shape microclimates. Trees in higher altitudes have been observed to adapt their branching to allow for light to pass through subsequent canopy layers.  
*Canopy structure and topography jointly constrain the microclimate of human-modified tropical landscapes. Global change biology, 24(11), pp.5243-5258] .See *Jucker et al. 2018 - max height was highly correlated to both LAI and surface roughness
The strength of buffering also varies with respect to geographic and climatic factors.
* Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]
* [@davis_microclimatic_2019]--Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperatures [@davis_microclimatic_2019]
However vertical climate gradients within forests are known to be steeper than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone. 
Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
For instance, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016). Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). 
At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.

@@

**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

* Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
- (Cavaleri et al. 2008) 
- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues. I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
- Elsa: see also H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009). -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.

@@

**Within a species, leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody ~~and herbaceous~~ species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. 
Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).

@@

Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient.

>>>>

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. Foliage density mediates light regime, where denser upper canopies exposed to greater PAR values limit light to canopy interiors, lower canopy layers, and the understory. 
This light gradient is more pronounced under broadleaf forests than needleleaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b].

@@

The strength of this buffering varies across forests, and increases with canopy cover, leaf area index ($LAI$) and water availability [@davis_microclimatic_2019]. Diurnally, tropical, temperate broadleaf and non-pine conifer forests offset temperatures positively and negatively, by maintaining cooler, lower daytime maximum temperature and vapor pressure deficit ($VPD$), and warmer higher nighttime minimum temperature in the understory, respectively, relative to pine and boreal forests and compared to nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, Defrenne et al. 2021, @davis_microclimatic_2019, @zellweger_seasonal_2019]. In comparison to neighboring tree crop agricultural plantations, intact tropical forests also maintain lower peak daytime $T_{air}$ and $VPD$, attesting to increased buffering with canopy cover [S. R. Hardwick et al. 2015, Junker et al. 2018]. Pine forests, on the other hand, that are open-forests with lower $LAI$ and sparse stands have lower buffering capacity and greater light permeability through canopies, thus maintain higher daytime maximum temperature and light availability in the understory, that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999]. These pine forests have also been observed to maintain warmer nighttime understory temperatures than above canopy, due to radiative heat loss from the soil, caused by daytime solar radiation [@vonarxSpatiotemporalEffectsForest2012a]. Canopy roughness also affects buffering, turbulent air flow and the canopy boundary layer. As a result, taller trees up to 20m in an an old-growth Borneo forest (which coincided with forests of higher LAI) have shown to lower mean and max of VPD and temperature. However, beyond a certain height, temperature and VPD offset plateaus [Junker et al. 2018]. At a species level, broadleaf trees with higher shade casting ability also contribute to greater temperature buffering [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a].

@@

With higher altitude and topography, forest light environment varies along with plant growth habits, therefore shapes microclimate conditions. *(do higher altitude forests have lower buffering capacity? due to open canopies?)* In a European silver birch forest, trees in higher altitude maintained open canopies with allowing more light through subsequent canopy layers, resulting in similar leaf traits across canopy positions, than in lower altitude trees [@rajsnerovaMorphologicalBiochemicalPhysiological2015]. Increased light permeability in higher altitude European sparse pine canopies, has also shown to increase understory light conditions along with higher maximum temperatures than lower altitude canopies, thus potentially having implications for seedling recruitment as temperatures rise [@vonarxSpatiotemporalEffectsForest2012a]. Vertical climate gradients are observed to be steeper within forests than those driven by elevation and latitude [@nakamuraForestsTheirCanopies2017; @martinCovarianceSunShade2020a]*what does this mean? can combine above sentences into one(N.V note)* Tree height in combination with topography modifies microclimate conditions. In a Boreneo forest, greater canopy height had a stronger influence on buffering temperature on east facing, wind-exposed slopes than on west-facing slopes.

The strength of buffering also varies with respect to geographic and climatic factors. Distance to coast, topographic position, elevation [@zellweger_seasonal_2019]. Proximity to riparian areas tends to sustain buffering, protecting understory temperatures even during increasing temperature extremes [@davis_microclimatic_2019]

@@

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone.
In open-canopy pine forests, light permeating through canopy layers can influence understory leaf temperatures, such that even with understories maintaing $T_{air}$ lower than above canopy $T_{air}$, lower canopy $T_{leaf}$ can be substantially higher than above canopy, even in moist conditions [@mullerEvidenceEfficientNonevaporative2021, @martin_boundary_1999]  
Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
For instance, during the dry season in a semi-decidudous tropical moist forest in Panama [@rey-sanchez_spatial_2016]. Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves [@rey-sanchez_spatial_2016]. 
At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.

@@

**Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

* Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
- (Cavaleri et al. 2008) 
- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues. I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
- Elsa: see also H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009). -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.
- Beyond a certain height, height becomes more important than light in determing traits [Ambrose, Sillet, Cavaleri]

@@

**Within a species, leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [@rozendaalPlasticityLeafTraits2006].**
After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody species [@keenanGlobalLeafTrait2016; @niinemetsLeafAgeDependent2016a]. 
A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR and the ratio of red to far red radiation intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [@casalPhotoreceptorSignalingNetworks2013]. 
Degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; @sackHowStrongIntracanopy2006], ultimately influencing leaf structural, chemical and physiological traits as well [@keenanGlobalLeafTrait2016].

@@

- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015), however stomatal conductance limitation is also greater, and earlier with canopy height [Sanches et al. 2009] 
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).
-understory leaves open their stomata wider than upper canopy leaves [Kosugi et al. 2021]

@@

Interaction between degrees of shade x ontogeny x height with rising temperature needs more scientific attention because a huge part of understory vegetation lies within variations of light and shade spatio-temporal gradient. Leaf traits vary more on the basis on light environment than ontogenetically [REF]
